---
title: "In-class_Ex05"
author: "Abinav Shajil"
format: html
date: "September 26, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

## Installing and Loading the R Packages

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## Import Shapefile into R Environment

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

## Import CSV File into R Environment

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```
## Performing Relational Join

```{r}
hunan_GDPPC <- hunan %>%
  left_join(hunan2012) %>%
  select(1:4, 7, 15, 33)
```

## Global Measures of Spatial Association
### Step 1: Deriving Queen's contiguity weights: sfdep methods

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

### Step 2: Computing Global Moran'I
```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```

##Performing Global Moran'I permutation test

```{r}
set.seed(!234) 

global_moran_perm (wm_q$GDPPC, 
                   wm_q$nb,
                   wm_q$wt,
                   nsim = 99)
```

## Computing local Moran's I


```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(local_moran)
```

## Plotting LISA map

```{r}
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)
tmap_mode("plot")
tm_shape(lisa) + 
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(lisa_sig) + 
  tm_polygons(fill = "mean", 
              fill.legend = tm_legend(
                text.size = 1.15,
                width = 6)) +
  tm_borders(fill_alpha = 0.4) + 
  tm_layout(legend.title.size = 1.25)
```

## Computing Local Gi* Statistics

```{r}
wm_idw <- hunan_GDPPC %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wts = st_inverse_distance(nb, 
                              geometry, 
                              scale = 1,
                              alpha = 1),
         .before = 1)
```

```{r}
HCSA <- wm_idw %>% 
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wts, nsim = 99),
         .before = 1) %>%
  unnest(local_Gi)
```

## Visualising Gi*

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(fill_alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_title(
    "local G*i for GDP per capita")
```
## Visualising p-value of local Gi*

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons(
    fill = "p_sim",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.001, 0.01, 0.05, 1.0),
      labels = c("p < 0.001", "p < 0.01", "p < 0.05", "Not Significant"),
      values = c("#de2d26", "#fc9272", "#fee0d2", "#e0e0e0")),
    fill.legend = tm_legend(
      title = "p-values")
    ) + 
  tm_title("p-values of G*i for GDP per capita") +
  tm_borders(fill_alpha = 0.5)
```

## Visualising hot spot and cold spot areas

```{r}
HCSA_sig <- HCSA  %>%
  filter(p_sim < 0.05)
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(HCSA_sig) +
  tm_polygons(fill = "cluster") + 
  tm_borders(fill_alpha = 0.4) +
  tm_title("Hot and Cold Spots map for GDP per capita")
```
