---
title: "Geospatial Analytics for Public Safety"
subtitle: "Take-home Exercise 1: Forest-Fire Events in Ketapang Regency, West Kalimantan"
date: "today"
date-modified: last-modified
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
toc: true
toc-location: right
---

```{r}
#| label: load-packages

pacman::p_load(sf, tidyverse, spatstat, tmap, terra)
```

## Introduction and Study Context

## Research Question and Objectives

**Research Question:** Where are forest-fire events concentrated within Ketapang Regency, and how does their spatial distribution and interaction change across the observed 2026 period (particularly as the dry season progresses)?

## Data and Study Design

**Study area:** Ketapang Regency, West Kalimantan, Indonesia.

**Core datasets:**

- Forest-fire event data: NASA FIRMS, VIIRS S-NPP (SV-C2) and VIIRS NOAA-20 (J1V-C2), bounding box `109.8, -3.2, 111.3, 0.0`, observation period 1 Jan 2026 – 12 Sept 2026, Shapefile format.
- Administrative boundary data: indonesia-geospasial.com, "Batas Administrasi Kabupaten/Kota Tahun 2019" (Indonesia regency boundaries), filtered to Ketapang Regency.

## Data Preparation and Quality Assessment

### Data Import

fire_archive_sv \<- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_archive_SV-C2_804529.shp")

fire_nrt_sv \<- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_nrt_SV-C2_804529.shp")

```{r}
#| label: import-fire-sv

fire_archive_sv <- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_archive_SV-C2_804529.shp")
fire_nrt_sv     <- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_nrt_SV-C2_804529.shp")
```

```{r}
#| label: import-fire-j1v

fire_archive_j1v <- st_read("data/geospatial/DL_FIRE_J1V-C2_804528/fire_archive_J1V-C2_804528.shp")
fire_nrt_j1v     <- st_read("data/geospatial/DL_FIRE_J1V-C2_804528/fire_nrt_J1V-C2_804528.shp")
```

```{r}
#| label: import-boundary

boundary_id <- st_read("data/geospatial/BATAS KABUPATEN KOTA DESEMBER 2019 DUKCAPIL/BATAS KABUPATEN KOTA DESEMBER 2019 DUKCAPIL.shp")
names(boundary_id)
```

```{r}
#| label: inspect-fire-columns

names(fire_archive_sv)
names(fire_nrt_sv)
```

```{r}
#| label: check-ketapang-name

boundary_id$KAB_KOTA[grepl("ketapang", boundary_id$KAB_KOTA, ignore.case = TRUE)]
```

```{r}
#| label: combine-fire-data

# Add source tags before combining, since we're losing archive/nrt 
# and satellite distinction once merged (relevant for data-quality discussion)
fire_archive_sv <- fire_archive_sv %>% mutate(source = "archive", satellite = "SV")
fire_nrt_sv     <- fire_nrt_sv     %>% mutate(source = "nrt", satellite = "SV", TYPE = NA)

fire_archive_j1v <- fire_archive_j1v %>% mutate(source = "archive", satellite = "J1V")
fire_nrt_j1v     <- fire_nrt_j1v     %>% mutate(source = "nrt", satellite = "J1V", TYPE = NA)

# Combine all four into one fire event dataset
fire_all <- bind_rows(fire_archive_sv, fire_nrt_sv, fire_archive_j1v, fire_nrt_j1v)

nrow(fire_all)
names(fire_all)
summary(fire_all$ACQ_DATE)
```

```{r}
#| label: filter-boundary-ketapang

# Filter boundary to Ketapang Regency only
ketapang_boundary <- boundary_id %>% 
  filter(KAB_KOTA == "KETAPANG")

# Transform to projected CRS (UTM Zone 49S) - needed for accurate distance-based 
# analysis later (spatstat requires planar coordinates, not lat/long degrees)
ketapang_boundary <- st_transform(ketapang_boundary, crs = 32749)

st_geometry_type(ketapang_boundary)
st_crs(ketapang_boundary)$epsg
st_area(ketapang_boundary)
```

```{r}
#| label: prep-and-filter-fire-points

# Convert ACQ_DATE from character to proper Date type
fire_all <- fire_all %>% 
  mutate(ACQ_DATE = as.Date(ACQ_DATE))

# Reproject fire points to match boundary CRS (UTM Zone 49S)
fire_all <- st_transform(fire_all, crs = 32749)

# Spatially filter to points falling within Ketapang boundary
fire_ketapang <- st_filter(fire_all, ketapang_boundary)

nrow(fire_ketapang)
class(fire_all$ACQ_DATE)
range(fire_ketapang$ACQ_DATE)
```

```{r}
#| label: data-quality-checks

# Check for duplicate observations (same lat/long/date/time)
n_duplicates <- fire_ketapang %>% 
  st_drop_geometry() %>% 
  duplicated() %>% 
  sum()

# Check for missing/invalid geometries
n_empty_geom <- sum(st_is_empty(fire_ketapang))

# Check CONFIDENCE field - VIIRS uses categorical low/nominal/high
table(fire_ketapang$CONFIDENCE, useNA = "always")

# Check for any NA coordinates (shouldn't happen post-import, but verify)
n_na_coords <- sum(is.na(st_coordinates(fire_ketapang)))

n_duplicates
n_empty_geom
n_na_coords
```

```{r}
#| label: filter-low-confidence

fire_ketapang_clean <- fire_ketapang %>% 
  filter(CONFIDENCE != "l")

nrow(fire_ketapang_clean)
table(fire_ketapang_clean$CONFIDENCE, useNA = "always")
```
