---
title: "Geospatial Analytics for Public Safety"
subtitle: "Take-home Exercise 1: Forest-Fire Events in Ketapang Regency, West Kalimantan"
date: "today"
date-modified: last-modified
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
toc: true
toc-location: right
---

```{r}
#| label: load-packages

pacman::p_load(sf, tidyverse, spatstat, tmap, terra)
```

## Introduction and Study Context

## Research Question and Objectives

**Research Question:** Where are forest-fire events concentrated within Ketapang Regency, and how does their spatial distribution and interaction change across the observed 2026 period (particularly as the dry season progresses)?

## Data and Study Design

**Study area:** Ketapang Regency, West Kalimantan, Indonesia.

**Core datasets:**

- Forest-fire event data: NASA FIRMS, VIIRS S-NPP (SV-C2) and VIIRS NOAA-20 (J1V-C2), bounding box `109.8, -3.2, 111.3, 0.0`, observation period 1 Jan 2026 – 12 Sept 2026, Shapefile format.
- Administrative boundary data: indonesia-geospasial.com, "Batas Administrasi Kabupaten/Kota Tahun 2019" (Indonesia regency boundaries), filtered to Ketapang Regency.

## Data Preparation and Quality Assessment

### Data Import

fire_archive_sv \<- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_archive_SV-C2_804529.shp")

fire_nrt_sv \<- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_nrt_SV-C2_804529.shp")

```{r}
#| label: import-fire-sv

fire_archive_sv <- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_archive_SV-C2_804529.shp")
fire_nrt_sv     <- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_nrt_SV-C2_804529.shp")
```

```{r}
#| label: import-fire-j1v

fire_archive_j1v <- st_read("data/geospatial/DL_FIRE_J1V-C2_804528/fire_archive_J1V-C2_804528.shp")
fire_nrt_j1v     <- st_read("data/geospatial/DL_FIRE_J1V-C2_804528/fire_nrt_J1V-C2_804528.shp")
```

```{r}
#| label: import-boundary

boundary_id <- st_read("data/geospatial/BATAS KABUPATEN KOTA DESEMBER 2019 DUKCAPIL/BATAS KABUPATEN KOTA DESEMBER 2019 DUKCAPIL.shp")
names(boundary_id)
```

```{r}
#| label: inspect-fire-columns

names(fire_archive_sv)
names(fire_nrt_sv)
```

```{r}
#| label: check-ketapang-name

boundary_id$KAB_KOTA[grepl("ketapang", boundary_id$KAB_KOTA, ignore.case = TRUE)]
```
