---
title: "Geospatial Analytics for Public Safety"
subtitle: "Take-home Exercise 1: Forest-Fire Events in Ketapang Regency, West Kalimantan"
date: "today"
date-modified: last-modified
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
toc: true
toc-location: right
---

```{r}
#| label: load-packages

pacman::p_load(sf, tidyverse, spatstat, tmap, terra)
```

## Introduction and Study Context

## Research Question and Objectives

**Research Question:** Where are forest-fire events concentrated within Ketapang Regency, and how does their spatial distribution and interaction change across the observed 2026 period (particularly as the dry season progresses)?

## Data and Study Design

**Study area:** Ketapang Regency, West Kalimantan, Indonesia.

**Core datasets:**

- Forest-fire event data: NASA FIRMS, VIIRS S-NPP (SV-C2) and VIIRS NOAA-20 (J1V-C2), bounding box `109.8, -3.2, 111.3, 0.0`, observation period 1 Jan 2026 – 12 Sept 2026, Shapefile format.
- Administrative boundary data: indonesia-geospasial.com, "Batas Administrasi Kabupaten/Kota Tahun 2019" (Indonesia regency boundaries), filtered to Ketapang Regency.

## Data Preparation and Quality Assessment

### Data Import

fire_archive_sv \<- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_archive_SV-C2_804529.shp")

fire_nrt_sv \<- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_nrt_SV-C2_804529.shp")

```{r}
#| label: import-fire-sv

fire_archive_sv <- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_archive_SV-C2_804529.shp")
fire_nrt_sv     <- st_read("data/geospatial/DL_FIRE_SV-C2_804529/fire_nrt_SV-C2_804529.shp")
```

```{r}
#| label: import-fire-j1v

fire_archive_j1v <- st_read("data/geospatial/DL_FIRE_J1V-C2_804528/fire_archive_J1V-C2_804528.shp")
fire_nrt_j1v     <- st_read("data/geospatial/DL_FIRE_J1V-C2_804528/fire_nrt_J1V-C2_804528.shp")
```

```{r}
#| label: import-boundary

boundary_id <- st_read("data/geospatial/BATAS KABUPATEN KOTA DESEMBER 2019 DUKCAPIL/BATAS KABUPATEN KOTA DESEMBER 2019 DUKCAPIL.shp")
names(boundary_id)
```

```{r}
#| label: inspect-fire-columns

names(fire_archive_sv)
names(fire_nrt_sv)
```

```{r}
#| label: check-ketapang-name

boundary_id$KAB_KOTA[grepl("ketapang", boundary_id$KAB_KOTA, ignore.case = TRUE)]
```

```{r}
#| label: combine-fire-data

# Add source tags before combining, since we're losing archive/nrt 
# and satellite distinction once merged (relevant for data-quality discussion)
fire_archive_sv <- fire_archive_sv %>% mutate(source = "archive", satellite = "SV")
fire_nrt_sv     <- fire_nrt_sv     %>% mutate(source = "nrt", satellite = "SV", TYPE = NA)

fire_archive_j1v <- fire_archive_j1v %>% mutate(source = "archive", satellite = "J1V")
fire_nrt_j1v     <- fire_nrt_j1v     %>% mutate(source = "nrt", satellite = "J1V", TYPE = NA)

# Combine all four into one fire event dataset
fire_all <- bind_rows(fire_archive_sv, fire_nrt_sv, fire_archive_j1v, fire_nrt_j1v)

nrow(fire_all)
names(fire_all)
summary(fire_all$ACQ_DATE)
```

```{r}
#| label: filter-boundary-ketapang

# Filter boundary to Ketapang Regency only
ketapang_boundary <- boundary_id %>% 
  filter(KAB_KOTA == "KETAPANG")

# Transform to projected CRS (UTM Zone 49S) - needed for accurate distance-based 
# analysis later (spatstat requires planar coordinates, not lat/long degrees)
ketapang_boundary <- st_transform(ketapang_boundary, crs = 32749)

st_geometry_type(ketapang_boundary)
st_crs(ketapang_boundary)$epsg
st_area(ketapang_boundary)
```

```{r}
#| label: prep-and-filter-fire-points

# Convert ACQ_DATE from character to proper Date type
fire_all <- fire_all %>% 
  mutate(ACQ_DATE = as.Date(ACQ_DATE))

# Reproject fire points to match boundary CRS (UTM Zone 49S)
fire_all <- st_transform(fire_all, crs = 32749)

# Spatially filter to points falling within Ketapang boundary
fire_ketapang <- st_filter(fire_all, ketapang_boundary)

nrow(fire_ketapang)
class(fire_all$ACQ_DATE)
range(fire_ketapang$ACQ_DATE)
```

```{r}
#| label: data-quality-checks

# Check for duplicate observations (same lat/long/date/time)
n_duplicates <- fire_ketapang %>% 
  st_drop_geometry() %>% 
  duplicated() %>% 
  sum()

# Check for missing/invalid geometries
n_empty_geom <- sum(st_is_empty(fire_ketapang))

# Check CONFIDENCE field - VIIRS uses categorical low/nominal/high
table(fire_ketapang$CONFIDENCE, useNA = "always")

# Check for any NA coordinates (shouldn't happen post-import, but verify)
n_na_coords <- sum(is.na(st_coordinates(fire_ketapang)))

n_duplicates
n_empty_geom
n_na_coords
```

```{r}
#| label: filter-low-confidence

fire_ketapang_clean <- fire_ketapang %>% 
  filter(CONFIDENCE != "l")

nrow(fire_ketapang_clean)
table(fire_ketapang_clean$CONFIDENCE, useNA = "always")
```

## First-Order Spatial Point Pattern Analysis

### Preparing the Point Pattern Object

```{r}
#| label: create-ppp-object

# Convert Ketapang boundary to an owin (observation window) for spatstat
ketapang_owin <- as.owin(ketapang_boundary)

# Extract coordinates from the cleaned fire points
fire_coords <- st_coordinates(fire_ketapang_clean)

# Create the ppp (planar point pattern) object
fire_ppp <- ppp(x = fire_coords[,1], 
                 y = fire_coords[,2], 
                 window = ketapang_owin)

# Targeted checks instead of full summary()
n_dupes_ppp <- sum(duplicated(fire_ppp))
n_points_final <- fire_ppp$n
n_points_original <- nrow(fire_coords)

cat("Points in ppp object:", n_points_final, "\n")
cat("Points before conversion:", n_points_original, "\n")
cat("Points dropped (outside window):", n_points_original - n_points_final, "\n")
cat("Duplicated coordinate pairs:", n_dupes_ppp, "\n")
```

### Kernel Density Estimation

```{r}
#| label: kde-fire-intensity

# Kernel density estimation using diggle bandwidth selection method
# (data-driven, avoids arbitrarily picking a bandwidth)
kde_fire <- density(fire_ppp, sigma = bw.diggle, edge = TRUE)

plot(kde_fire, main = "Kernel Density Estimate of Fire Events in Ketapang")
plot(ketapang_boundary$geometry, add = TRUE, border = "white")
```

```{r}
#| label: check-bandwidth

bw.diggle(fire_ppp)
```

```{r}
#| label: compare-bandwidths

bw_diggle <- bw.diggle(fire_ppp)
bw_ppl    <- bw.ppl(fire_ppp)
bw_scott  <- bw.scott(fire_ppp)

cat("Diggle:", bw_diggle, "\n")
cat("PPL:", bw_ppl, "\n")
cat("Scott (x):", bw_scott[1], "\n")
cat("Scott (y):", bw_scott[2], "\n")
```

```{r}
#| label: kde-final-and-robustness

# Primary KDE using Diggle's bandwidth (data-driven, suited to clustered patterns)
kde_diggle <- density(fire_ppp, sigma = bw_diggle, edge = TRUE)

# Robustness check: KDE using Scott's rule (much larger bandwidth)
kde_scott <- density(fire_ppp, sigma = bw_scott, edge = TRUE)

par(mfrow = c(1,2))
plot(kde_diggle, main = "KDE (Diggle bandwidth)")
plot(ketapang_boundary$geometry, add = TRUE, border = "white")
plot(kde_scott, main = "KDE (Scott bandwidth)")
plot(ketapang_boundary$geometry, add = TRUE, border = "white")
par(mfrow = c(1,1))
```

### Temporal Variation in Fire Intensity (Phase-Based)

```{r}
#| label: phase-kde-prep

# Define fire phase based on documented 2026 Kalimantan wildfire escalation timeline
fire_ketapang_clean <- fire_ketapang_clean %>% 
  mutate(phase = case_when(
    ACQ_DATE < as.Date("2026-07-01") ~ "Pre-escalation (Jan-Jun)",
    ACQ_DATE < as.Date("2026-08-01") ~ "Onset (Jul)",
    TRUE ~ "Crisis (Aug-Sep)"
  ),
  phase = factor(phase, levels = c("Pre-escalation (Jan-Jun)", "Onset (Jul)", "Crisis (Aug-Sep)")))

table(fire_ketapang_clean$phase)
```

```{r}
#| label: phase-kde

# Split ppp object by phase (factor levels set in phase-kde-prep control order)
fire_ppp_phase <- fire_ppp
marks(fire_ppp_phase) <- fire_ketapang_clean$phase

fire_split_phase <- split(fire_ppp_phase)

# KDE per phase, same bandwidth (bw_diggle) for comparability across panels
kde_phase <- lapply(fire_split_phase, function(p) density(p, sigma = bw_diggle, edge = TRUE))

par(mfrow = c(1,3), mar = c(1,1,2,1))
for (ph in names(kde_phase)) {
  plot(kde_phase[[ph]], main = ph)
  plot(ketapang_boundary$geometry, add = TRUE, border = "white")
}
par(mfrow = c(1,1))
```

```{r}
#| label: quadrat-test-csr

# Quadrat count test for Complete Spatial Randomness (CSR)
# Using a 5x5 grid - fine enough to detect structure, coarse enough for stable counts
qcount <- quadratcount(fire_ppp, nx = 5, ny = 5)
qtest <- quadrat.test(fire_ppp, nx = 5, ny = 5)

qtest
```

### Nearest-Neighbour Distance Functions (G and F)

```{r}
#| label: g-function-observed

set.seed(1234)

g_fire <- Gest(fire_ppp)
plot(g_fire, main = "G-function: Observed Nearest-Neighbour Distances")
```

```{r}
#| label: f-function-observed

set.seed(1234)

f_fire <- Fest(fire_ppp)
plot(f_fire, main = "F-function: Observed Empty Space Distances")
```

```{r}
#| label: g-function-envelope

set.seed(1234)

g_envelope <- envelope(fire_ppp, Gest, nsim = 99, rank = 1)
plot(g_envelope, main = "G-function with CSR Envelope (99 simulations)")
```

```{r}
#| label: f-function-envelope

set.seed(1234)

f_envelope <- envelope(fire_ppp, Fest, nsim = 99, rank = 1)
plot(f_envelope, main = "F-function with CSR Envelope (99 simulations)")
```

### Second-Order Interaction (K and L Functions)

```{r}
#| label: k-function-observed

set.seed(1234)

k_fire <- Kest(fire_ppp, rmax = 5000, correction = "border")
plot(k_fire, main = "K-function: Observed (rmax = 5km)")
```

```{r}
#| label: k-function-envelope

set.seed(1234)

k_envelope <- envelope(fire_ppp, Kest, rmax = 5000, correction = "border", nsim = 99, rank = 1)
plot(k_envelope, main = "K-function with CSR Envelope (99 simulations, rmax = 5km)")
```

```{r}
#| label: l-function-envelope

set.seed(1234)

l_envelope <- envelope(fire_ppp, Lest, rmax = 5000, correction = "border", nsim = 99, rank = 1)
plot(l_envelope, . - r ~ r, main = "L-function with CSR Envelope (centred, rmax = 5km)")
```

### Value-Added Analysis: Spatio-Temporal Interaction Comparison

```{r}
#| label: spatiotemporal-k-observed

set.seed(1234)

# Reuse phase split from earlier (fire_split_phase)
k_by_phase <- lapply(fire_split_phase, function(p) Kest(p, rmax = 5000, correction = "border"))

plot(k_by_phase[[1]], main = "K-function by Phase", col = "blue", lty = 1)
plot(k_by_phase[[2]], add = TRUE, col = "orange", lty = 1)
plot(k_by_phase[[3]], add = TRUE, col = "red", lty = 1)
legend("topleft", legend = names(k_by_phase), col = c("blue","orange","red"), lty = 1)
```

```{r}
#| label: spatiotemporal-k-envelopes

set.seed(1234)

k_envelope_by_phase <- lapply(fire_split_phase, function(p) 
  envelope(p, Kest, rmax = 5000, correction = "border", nsim = 99, rank = 1))

par(mfrow = c(1,3), mar = c(4,4,2,1))
for (ph in names(k_envelope_by_phase)) {
  plot(k_envelope_by_phase[[ph]], main = ph)
}
par(mfrow = c(1,1))
```
